Parsing Postfix log files

### 2009-03-09 doc/logparser-architecture.tex
Corrections and changes after comments from Carl.

For each input the framework tries each rule in turn until it finds a rule
that recognises the input, then invokes the action specified by that rule.
If the input is not recognised by any of the rules, the framework issues a
warning; the framework can continue parsing after this, although for some
inputs stopping immediately may be preferable.

@@

Separating the rules from the actions and framework makes it possible to
parse new inputs without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to recognise the inputs is
trivial in comparison to understanding an entire parser, identifying the
correct location to change, and making the appropriate changes.  Bear in
mind that changes to a parser must be made without adversely affecting
existing parsing, including any edge cases that are not immediately obvious
--- \sectionref{yet-more-aborted-delivery-attempts} describes a
complication that occurs only four times in \numberOFlogFILES{} log files.
Requiring changes to a parser's source code also complicates upgrades of
the parser, because the changes must be preserved during the upgrade, and
they may clash with changes made by the developer.  This architecture
allows the user to add new rules without having to edit a parser, unless
the new inputs cannot be processed by the existing actions.  If the new
inputs do require new functionality, new actions can be added to the parser
without having to modify existing actions; only in the rare event that the
new actions need to cooperate with existing actions will more extensive
changes be required.

@@

@@

    \item [Save and load state]  The architecture can save its current
        state, i.e.\ the shared storage it provides for actions, and reload
        it later, so that information is not lost between parsing runs.

@@

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejecting a delivery attempt.  There may be many
input variants within one input category; in general each action will
handle one input category, with each rule recognising one input variant.
It is anticipated that parsers based on this architecture will have a high
ratio of rules to actions, with the aim of having simple rules and tightly
focused actions.  An action may need to process different input variants in
slightly different ways, but large changes in processing indicate the need
for a new action and a new category of input; if an action becomes overly
complicated it starts to turn into a monolithic parser, with too much logic
contained in a single procedure.

@@

Actions may return a modified input line to the framework that will be
parsed as if read from the input stream, allowing for a simplified version
of cascaded parsing~\cite{cascaded-parsing}.  This powerful facility allows
several rules and actions to parse a single input, potentially simplifying
both rules and actions.  A simple example is to have one rule and action
removing comments from input lines, so that other rules and actions do not
have to handle comments at all; obviously if comment characters can be
escaped or embedded in quoted strings the implementation needs to be more
complex, but for some inputs this can greatly simplify subsequent parsing.
Actions do not need to be specially registered with the framework or be
declared in a particular way to use cascaded parsing: actions that do not
use cascaded parsing will return nothing, those that do will simply return
a string to be parsed.

@@

The architecture is line oriented at present: there is no facility for
rules to consume more input or push unused input back onto the input
stream.  This was not a deliberate design decision, rather a consequence of
the line oriented nature of Postfix log files; more flexible approaches
could be pursued.  Classifying inputs using the rules is simple: the first
rule to recognise the input determines the action that will be invoked;
there is no backtracking to try alternate rules; no attempt is made to pick
a \textit{best\/} rule.  \sectionref{attaching conditions to rules}
contains an example which requires that the rules are used in a specific
order to correctly parse the input, so a mechanism is needed to allow the
author of the rules to specify that ordering.  The framework should sort
the rules according to a priority attribute specified by the rule author,
allowing fine-grained control over the order that rules are used in.  The
priority attribute may be implemented as a number, or as a range of values,
e.g.\ low, medium, and high, or in a different fashion entirely if it suits
the implementation.  Rule ordering for efficiency is a separate topic that
is covered in \sectionref{rule ordering for efficiency}; overlapping rules
are discussed in \sectionref{overlapping rules in architecture}.

@@

Conditions that examine the input will be the easiest to understand,
because they can be understood in isolation; they do not depend on
variables set by actions or other rules.  Conditions that examine the input
can be complex if required, but simple conditions can be quite useful too,
e.g.\ every Postfix log \empty{}line contains the name of the Postfix
component that produced it, so every rule used in \parsername{} has a
condition specifying the component whose log \empty{}lines it recognises,
reducing the number of rules that will be used when classifying an input
(see \sectionref{rules in implementation} for details) and increasing the
chance that the input will be correctly recognised..

>>>>

For each input, the framework tries each rule in turn until it finds a rule
that recognises the input, then invokes the action specified by that rule.
If the input is not recognised by any of the rules, the framework issues a
warning; the framework can continue parsing after this, although for some
inputs stopping immediately may be preferable.

@@

Separating the rules from the actions and framework makes it possible to
parse new inputs without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to recognise the inputs is
trivial in comparison to understanding an entire parser, identifying the
correct location to change, and making the appropriate changes.  Additions
to a parser must be made without adversely affecting existing parsing,
including any edge cases that are not immediately obvious; an edge case
that occurs only four times in \numberOFlogFILES{} log files is described
in Yet More Aborted Delivery Attempts
(\sectionref{yet-more-aborted-delivery-attempts}).  More intrusive changes
are more likely to introduce a bug, so reducing the extent of the changes
is important.  Requiring changes to a parser's source code also complicates
upgrades of the parser, because the changes must be preserved during the
upgrade, and they may clash with changes made by the developer.  This
architecture allows the user to add new rules without having to edit a
parser, unless the new inputs cannot be processed by the existing actions.
If the new inputs do require new functionality, new actions can be added to
the parser without having to modify existing actions; only in the rare
event that the new actions need to cooperate with existing actions will
more extensive changes be required.

@@

    \item [Convert physical inputs to logical inputs] Each rule recognises
        a single input at a time: there is no facility for rules to consume
        more input or push unused input back onto the input stream,
        although actions may use cascaded parsing (\sectionref{cascaded
        parsing}) to effectively push input back onto the input stream.  A
        physical input (i.e.\ a single line read from the input stream) may
        contain multiple or partial logical inputs, and the framework must
        pre-process these to provide a logical input for the rules to
        recognise.  This simplifies the rules and actions considerably, at
        the cost of added complexity in the framework; during the design
        phase it was decided that it was easier to deal with the problem of
        parsing multiple or partial inputs once, rather than dealing with
        it in every rule and action.  This is easy in the case of Postfix
        log files because there is a one-to-one mapping between physical
        and logical inputs; mapping between physical and logical inputs may
        be more difficult for other types of input.  Some input types may
        require pre-processing equivalent to parsing the physical inputs;
        in those cases the framework should behave similarly to other
        architectures: present the entire input stream at once, and discard
        the portion recognised by the successful rule.

@@

    \item [Save and load state]  The architecture can save its current
        state, i.e.\ the shared storage it provides for actions, and reload
        it later, so that information is not lost between parsing runs.
        \parsername{} uses this functionality because mails may take some
        time to deliver and thus have their log lines split between log
        files; a compiler might store data structures it builds as it
        parses different files.

@@

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejecting a delivery attempt.  There may be many
input variants within one input category; in general each action will
handle one input category, with each rule recognising one input variant.
It is expected that parsers based on this architecture will have many
actions, and each action may be specified by many rules, with the aim of
having simple rules and tightly focused actions.  An action may need to
process different input variants in slightly different ways, but large
changes in processing indicate the need for a new action and a new category
of input; if an action becomes overly complicated it starts to turn into a
monolithic parser, with too much logic contained in a single procedure.

@@

Actions may return a modified input line to the framework that will be
parsed as if read from the input stream, allowing for a simplified version
of cascaded parsing~\cite{cascaded-parsing}.  This powerful facility allows
several rules and actions to parse a single input, potentially simplifying
both rules and actions.  A simple example is to have one rule and action
removing comments from input lines, so that other rules and actions do not
have to handle comments at all; obviously if comment characters can be
escaped or embedded in quoted strings the implementation needs to be more
complex, but for some inputs this can greatly simplify subsequent parsing.
A more complex example is evaluating simple arithmetic expressions, where
sub-expressions enclosed in parentheses must be evaluated first; cascaded
parsing can be used to parse and evaluate sub-expressions, substituting the
result into the original expression.  Actions do not need to be specially
registered with the framework or be declared in a particular way to use
cascaded parsing: actions that do not use cascaded parsing will return
nothing, those that do will simply return a string to be parsed.

XXX MAYBE MOVE THIS EXAMPLE TO AN APPENDIX\@?  DO I EVEN NEED IT\@?  ASK
CARL ABOUT IT\@.

@@

Classifying inputs using the rules is simple: the first
rule to recognise the input determines the action that will be invoked;
there is no backtracking to try alternate rules; no attempt is made to pick
a \textit{best\/} rule.  \sectionref{attaching conditions to rules}
contains an example which requires that the rules are used in a specific
order to correctly parse the input, so a mechanism is needed to allow the
author of the rules to specify that ordering.  The framework should sort
the rules according to a priority attribute specified by the rule author,
allowing fine-grained control over the order that rules are used in.  The
priority attribute may be implemented as a number, or as a range of values,
e.g.\ low, medium, and high, or in a different fashion entirely if it suits
the implementation.  Rule ordering for efficiency is a separate topic that
is covered in \sectionref{rule ordering for efficiency}; overlapping rules
are discussed in \sectionref{overlapping rules in architecture}.

@@

Conditions that examine the input will be the easiest to understand,
because they can be understood in isolation; they do not depend on
variables set by actions or other rules.  Conditions that examine the input
can be complex if required, but simple conditions can be quite useful too,
e.g.\ every Postfix log line contains the name of the Postfix component
that produced it, so every rule used in \parsername{} has a condition
specifying the component whose log lines it recognises, reducing the number
of rules that will be used when classifying an input (see \sectionref{rules
in implementation} for details) and increasing the chance that the input
will be correctly recognised.
